---
title: "Stat210 Final Project"
author: "Kevin Liu & Evan Zhong"
format: pdf
---
```{r read-data}
library(tidyverse)
library(dplyr)
library(tidymodels)
library(MASS)
library(nnet)
cbb <- read.csv("cbb.csv")
cbb <- subset(cbb, select = -W)
cbb <- cbb |> filter(!is.na(POSTSEASON)) |> 
              filter(POSTSEASON != "N/A" & !is.na(POSTSEASON))
#we only want to use the teams that make the tournament

```

```{r, message = F, warning = F}
cbbAug <- cbb |>
  mutate(POSTSEASON = case_when(
    POSTSEASON == "R68" ~ 68,
    POSTSEASON == "R64" ~ 64,
    POSTSEASON == "R32" ~ 32,
    POSTSEASON == "S16" ~ 16,
    POSTSEASON == "E8" ~ 8, 
    POSTSEASON == "F4" ~ 4,
    POSTSEASON == "2ND" ~ 2,
    POSTSEASON == "Champions" ~ 1
  ))
cbb <- cbb |>
  mutate(POSTSEASON = case_when(
    POSTSEASON == "R68" ~ 1,
    POSTSEASON == "R64" ~ 2,
    POSTSEASON == "R32" ~ 3,
    POSTSEASON == "S16" ~ 4,
    POSTSEASON == "E8" ~ 5,
    POSTSEASON == "F4" ~ 6,
    POSTSEASON == "2ND" ~ 7,
    POSTSEASON == "Champions" ~ 8
  ))
```

```{r}

cbbOrdinal <- polr(as.factor(POSTSEASON) ~ SEED + G + ADJOE + ADJDE + BARTHAG + EFG_O + EFG_D + TOR + TORD + ORB + DRB + FTR + FTRD + X2P_O + X2P_D + X3P_O + X3P_D + ADJ_T + WAB,  data = cbb)
summary(cbbOrdinal)
```

The coefficient corresponding to 8 seed is -0.22555. This means that when controlling for the other variables, if a team is an 8 seed as opposed to a 1 seed, then they are predicted to have e^-0.22555 (0.798077156) times the odds of making the next round of the NCAA tournament

The coefficient corresponding to adjusted offensive efficiency is 0.26490. This means that when controlling for the other variables, for team with one more adjusted offensive efficiency rating point compared to a team without that extra point, the team with the extra point is predicted to have e^-0.26490 (1.303) times the odds of making the next round of the NCAA tournament






```{r, message = F, warning = F}

cbbLinear <- lm(POSTSEASON ~ CONF + G + ADJOE + ADJDE + BARTHAG + EFG_O + EFG_D + TOR + TORD + ORB + DRB + FTR + FTRD + X2P_O + X2P_D + X3P_O + X3P_D + ADJ_T + WAB + SEED, data = cbbAug)
summary(cbbLinear)

augCbbLinear = augment(cbbLinear)

augCbbLinear |> 
  ggplot(aes(x = .fitted, y = .resid)) + 
  geom_point() +
  geom_hline(yintercept = 0, color = "blue") + 
  labs(x = "Fitted (Predicted) value", y = "Residuals",
       title = "Residual Model for Predicting Tournament Success (Multiple Negative Slopes)")

 augCbbLinear |> ggplot(aes(sample = .resid)) + 
  stat_qq() + 
  stat_qq_line() + 
  labs(x = "Theoretical quantiles", y = "Sample quantiles",
       title = "QQ Plot of Residuals Showing Minimal Deviation For Tournament Success",
       subtitle = "Is a linear line -- Can assume normality")
 
```
Assumptions needed for the linear model: Independence, Linearity, Constant Variance, and Normal Distribution

1. Independence
The dataset includes records of NCAA March Madness Tournament team performances (playoff runs) from 2013 to 2021. These observations--tournament runs—are NOT independent. Several factors contribute to this--for example, teams retaining players across seasons, powerhouse teams that show dominance throughout the years, the trading of players between teams and more. Thus, a team's performance in one tournament could influence its performance in future tournaments as well as other teams in the current tournament. Therefore, the assumption of independence is not met.

2. Linearity
  Linearity is not satisfied because the residuals are not symmetrically distributed along y = 0. The data points follow several very strong negative trends. This could be for many reasons, most likely because there are interaction effects. All in all though, the residual plot is not symmetric and therefore does not satisfy linearity.
  
3. Constant Variance
  Constant variance is not satisfied because the data set is not consistent/spreadout across all the predicted values. Although there is no signicant clumps, there are clear trends. Based on these features, we can conclude that the variance is not independent from the predictors

4. Normal Distribution
  Because the QQ plot follows the linear line, therefore we cannot assume normality.
  
# Interpretations
Holding all other variables constant, if a team is in Conference USA we'd expect them to place 17 higher than a team in the West Coast Conference

Holding all other variables constant, then, on average, for every additional point of adjusted offensive efficiency, we'd expect a team to place 2.53446 ranks higher

Holding all other variables constant, then, on average, for every additional point of adjusted defensive efficiency, we'd expect a team to place 3.13359 ranks higher

Hypothesis Test: 
Null Hypothesis, H0: there is not a linear relationship between 2-pt field goal percentage and how far a team makes it in the NCAA tournament.
Alternative Hypothesis, Ha: there is a linear relationship between 2-pt field goal percentage and how far a team makes it in the NCAA tournament.

Given that there is no relationship between 2-pt field goal percentage and how far a team makes it in the NCAA tournament, the probability of getting a slope of -5.41850 or less is 0.0187. Taking a 0.05 significance level, since the p-value is smaller than alpha (0.0187<0.05), there is convincing and sufficient evidence that there is a linear relationship between 2-pt field goal percentage and how far a team makes it in the NCAA tournament.

```{r}
cbbNullModel <- polr(factor(POSTSEASON)~ 1, data = cbb, Hess = TRUE)
logL_model <- logLik(cbbOrdinal)
logL_null <- logLik(cbbNullModel)

mcfadden_r_squared <- 1 - (as.numeric(logL_model) / as.numeric(logL_null))
print(mcfadden_r_squared)

```
Moderate Explanatory Power: A McFadden's R-squared value of about 0.25 suggests that the model has moderate explanatory power. While it's not particularly high (such as values closer to 0.5 or above), it's substantial enough to suggest that the model does provide valuable insights into the relationship between the conference (CONF) and the postseason outcomes (POSTSEASON).
ˆ




